Concurrent Programming with CXL

### 2020-06-14 book/paper.tex
Latest version

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}~\cite{courtois71}.
[RVR: or was it David Parnas who first introduced them?]
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The lock is used to protect access to those counters.
The program shows a process that executes in a loop.
Each time, it decides whether to read or write.
The critical section is spread between two labels:
readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
specification for two processes.

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
(aka spin-waiting):
processes spin in a loop until some desirable application-level condition is met.

The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:spinlocks}).
In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

But even when multiple processes each run on their own core, the solution
is inefficient.  After all, when the lock becomes available, the process
has to check other variables to see if it can continue.  In our case,
the process may have to read \texttt{nreaders} and/or \texttt{nwriters}
to decide if it can go on or has to try again, involving releasing
the lock and re-acquiring it.

A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
It uses two locks: \texttt{rwlock} is used by both readers and writers,
while \texttt{rlock} is only used by readers.
\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
or when a writer is at label \texttt{@wcs}.
\texttt{rlock} is used to protect the \texttt{nreaders} variable that
counts the number of readers in the critical section.

>>>>

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}~\cite{courtois71}.
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
We will call the lock the \emph{mutex} to distinguish it clearly from
the reader/writer lock that we are implementing.
The mutex is used to protect shared access to the counters.
The program shows a process that executes in a loop.
Each time, it decides whether to read or write.
The critical section is spread between two labels:
readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
specification for two processes.

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
(aka spin-waiting):
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}: they are in an infinite loop
in which they do not change the state.
The processes waiting for a reader/writer lock do change the state while
in the infinite loop: they acquire the mutex and release it.

In most operating systems and programming language runtimes,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
It uses two ordinary locks: \texttt{rwlock} is used by both readers and writers,
while \texttt{rlock} is only used by readers.
\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
or when a writer is at label \texttt{@wcs}.
\texttt{rlock} is used to protect the \texttt{nreaders} variable that
counts the number of readers in the critical section.
